Conservation and losses non-coding RNA associated loci in avian genomes

### 2014-08-07 paper/manuscript.tex
Updated manuscript to address reviewer comments.

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Conservation and losses of avian non-coding RNA loci}

\author{
Paul P.\ Gardner\correspondingauthor$^{1,2}$
\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Mario Fasold$^{5,6}$
\email{mario@bioinf.uni-leipzig.de},
Sarah W.\ Burge$^3$
\email{swb@ebi.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Stephanie Kehr$^5$
\email{steffi@bioinf.uni-leipzig.de},
Tammy E.\ Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter F.\ Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany.
\iid(6) ecSeq Bioinformatics, Brandvorwerkstr.43, D-04275 Leipzig, Germany.
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show that a number of
lncRNA-associated loci are conserved between birds and mammals,
including several intriguing cases where the reported mammalian lncRNA
function is not conserved in birds.  We also demonstrate extensive
conservation of classical ncRNAs (e.g., tRNAs) and more recently
discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
describe numerous ``losses'' of several RNA families, and attribute
these to genuine loss, divergence or missing data.  In particular, we
show that many of these losses are due to the challenges associated
with assembling Avian microchromosomes. These combined results illustrate the utility of
applying homology-based methods for annotating novel vertebrate
genomes.
\end{abstract}

@@

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the
recently published 45 avian genomes
\cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
provides an exciting opportunity to explore ncRNA conservation in
unprecedented detail.
%{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
%  CHROMOSOMES.}

@@

There is substantial gain and loss of lncRNAs and other ncRNA loci
over evolutionary time
\cite{Cabili:2011,Kutter:2012,Hoeppner:2012}. It is difficult to
assess how many of these ``gains'' and ``losses'' are due to limited
bioinformatic sequence alignment tools (these generally fail align
correctly below 60-50\% sequence identity
\cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or due to genuine
gains and losses. Nevertheless, sequence conservation, generally
speaking, provides useful evidence for gene and function conservation.

We have identified 66,879 loci in 48 avian genomes that share sequence
similarity with previously characterised ncRNAs and are conserved in
$>10\%$ of these avian genomes. These loci have been classified into
626 different families, the majority of which correspond to miRNAs and
snoRNAs (summarised in Table~\ref{table:1}). Out of necessity we have
selected the most charismatic families for further discussion. These
include the lncRNAs that appear to be conserved between Mammals and
Aves and the cases of apparent loss of genes that conserved in most
other Vertebrates.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus they
support a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1,
SOX2OT and ST7-OT3 lncRNAs have multiple, well conserved RNA-domains
(See Figure~\ref{fig:1}). The syntenic ordering of these seven lncRNAs
and the flanking genes are also preserved between the human and
chicken genomes (data not shown). We illustrate this in detail for the
HOTAIRM1 lncRNA (See Supplemental Figures~13\&14).

>>>>

%\title{Non-coding RNAs of bird genomes}
%\title{Wins and losses: Non-coding RNAs of bird genomes}
\title{Conservation and losses non-coding RNA associated loci in avian genomes}

\author{
Paul P.\ Gardner\correspondingauthor$^{1,2}$
\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterburmotifs/manuscript/y.ac.nz},
Mario Fasold$^{5,6}$
\email{mario@bioinf.uni-leipzig.de},
Sarah W.\ Burge$^3$
\email{swb@ebi.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Stephanie Kehr$^5$
\email{steffi@bioinf.uni-leipzig.de},
Tammy E.\ Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter F.\ Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany.
\iid(6) ecSeq Bioinformatics, Brandvorwerkstr.43, D-04275 Leipzig, Germany.
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNA loci in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We identify 34
lncRNA-associated loci that are conserved between birds and mammals
and validate 12 of these in chicken. These include several intriguing
cases where the reported mammalian lncRNA function is not conserved in
birds.  We also demonstrate extensive conservation of classical ncRNAs
(e.g., tRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and
miRNAs) in birds. Furthermore, we describe numerous ``losses'' of
several RNA families, and attribute these to genuine loss, divergence
or missing data.  In particular, we show that many of these losses are
due to the challenges associated with assembling Avian
microchromosomes. These combined results illustrate the utility of
applying homology-based methods for annotating novel vertebrate
genomes.
\end{abstract}

@@

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the
recently published 45 avian genomes
\cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
provides an exciting opportunity to explore conservation of genomic
loci that have been associated with ncRNAs in unprecedented detail.
%{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
%  CHROMOSOMES.}

@@

There is substantial gain and loss of lncRNAs and other ncRNA
associated loci over evolutionary time
\cite{Cabili:2011,Kutter:2012,Hoeppner:2012}. It is difficult to
assess how many of these ``gains'' and ``losses'' are due to limited
bioinformatic sequence alignment tools (these generally fail align
correctly below 60-50\% sequence identity
\cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or due to genuine
gains and losses or data missing from the current
genome assemblies. Nevertheless, sequence conservation, generally speaking,
provides useful evidence for gene and function conservation.

We have identified 66,879 loci in 48 avian genomes that share sequence
similarity with previously characterised ncRNAs and are conserved in
$>10\%$ of these avian genomes. These loci have been classified into
626 different families, the majority of which correspond to miRNAs and
snoRNAs (summarised in Table~\ref{table:1}). Out of necessity we have
selected a modest number of families for further discussion. These
include the lncRNAs that appear to be conserved between Mammals and
Aves and the cases of apparent loss of genes that conserved in most
other Vertebrates.

@@

When analysing the RNA-domain annotations it is striking that the
order (synteny) of many of the lncRNAs with multiple RNA-domains are
consistently preserved in the birds. The annotations of these domains
lie in the same genomic region, in the same order as in the mammalian
homologs. Thus they support a high degree of evolutionary conservation
for the entire lncRNA. In particular the HOXA11-AS1, PART1, PCA3,
RMST, Six3os1, SOX2OT and ST7-OT3 lncRNAs have multiple, well
conserved RNA-domains (See Figure~\ref{fig:1}). The syntenic ordering
of these seven lncRNAs and the flanking genes are also preserved
between the human and chicken genomes (data not shown). We illustrate
this in detail for the HOTAIRM1 lncRNA (See Supplemental
Figures~13\&14).
